Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-23 paper.tex
Updates

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
% Early in the SARS-CoV-2 pandemic there was insufficient
% genetic variation for meaningful recombination.
Early on in the COVID-19 pandemic, the levels of genetic diversity
were too low for the emergence of distinctive recombinant strains.
By late 2020, however, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur
and recombinant strains with onward transmission were detected at
low frequency~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of the SARS-CoV-2
virus.

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual process, subject
to biases [FIXME: have we a citation for how tedious this is? Also what
are the biases?]
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of characteristic mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
and as a result can only identify inter-lineage recombination events.
% specific to those motifs.
% % JK: I've left this out because I felt it broke the narrative up a bit, but
% % happy to put back in. (We're still citing the paper in a few places).
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
mutations.
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination), and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated ancestry.

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analyses, e.g., inference of
selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
no standard method for incorporating the effects of recombination into
phylogenetic analysis.
Ancestral Recombination Graphs (ARGs) are a means of describing such
network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
until recently lacked software support and sufficiently scalable
inference methods to be of practical use.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. The ``succinct tree sequence'' is an ARG data structure
which has led to significant computational advances across a range
of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
Wohns2022-th}, and the supporting \texttt{tskit} software library
is now widely used in population genetic applications.
The methods in \texttt{tskit} have been developed to support millions of
whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
to representing large SARS-CoV-2 genealogies,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}.
See Section~\ref{sec:args} for more details on ARGs and the succinct
tree sequence data structure.

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
Building on the open-source \texttt{tskit} library, the method infers
a joint model of mutation, genetic inheritance and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
First, likely attachment nodes in the current ARG
(or paths, if inferred to be a recombinant) are
chosen for each of the samples in the daily batch using the
Li and Stephens model (Section~\ref{sec:ls}). Then, since many samples
in a daily batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony heuristics to address issues
introduced by the inherent greediness of the strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of an accurate joint model of
mutation, genetic inheritance and recombination, will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development.
We work with a
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited, and
includes all 1.27 million sample genomes collected up to June 30, 2021
with a maximum delay between sampling and submission dates of 30 days
(see Section~\ref{sec:filtering_time_travellers}).
% The Long ARG is more sparsely sampled but
% includes data for an additional year (June 30, 2022).
% For the Long ARG, we use at most 1,000 randomly subsampled
% genomes per day from those meeting the maximum sampling delay criterion.
For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
day (again restricting the delay between sampling and submission to 30 days),
and include an additional year worth of samples (to June 30, 2022).

@@

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For example, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks~\citep{Kluyver2016-jupyter}
on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes, e.g.\ restricted by time or limited to close
relatives. We expect ARGs to be particularly tree-like
early in the pandemic, when co-infection was less likely.

@@

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the Nextstrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded Nextstrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the \texttt{sc2ts} ARGs to a backbone containing only samples present
the Nextstrain tree. This results in a small set of shared
samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
assigned to Pango recombinant lineages by Nextclade (see Section
\ref{sec:pango_x_lineages}).

@@

Figure~\ref{fig:cophylogeny} also reveals some notable differences between the
trees. Firstly, the \texttt{sc2ts} tree is generally less well resolved,
particularly in early 2020 when sampling density was much lower than
later in the pandemic.
Secondly, there are non-identical sample partitions
near the tips (e.g., in the Alpha clade).
Finally, and most noticeably, the branch lengths differ.
As discussed in Section~\ref{sec:node_dating}, the dating of nodes
other than samples in \texttt{sc2ts} is currently quite crude, and
it is likely that branches not ``anchored'' by samples will be
poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
these issues may be alleviated in future versions of \texttt{sc2ts}.

@@

\subsection{Mutational spectrum}
\label{sec:mutation_spectrum}
The ARGs inferred by \texttt{sc2ts} and represented using the \texttt{tskit}
library (Section~\ref{sec:args}) are a joint estimate of the processes
of genetic inheritance, recombination and mutation. Unlike most approaches to
phylogenetic analysis, mutations are included in the \texttt{tskit}
data model alongside
the topological representation of genetic inheritance.
This has many
advantages, for example allowing us to compute statistics of the observed
sequences efficiently~\citep{Kelleher2016-wk,Ralph2020-efficiently} and
to provide high-levels of data compression~\citep{Kelleher2019-ba}.
The same idea has recently been used to represent
SARS-CoV-2 data in UShER's ``mutation annotated tree''
format~\citep{Turakhia2021-ur}.

@@

Despite the presence of some artefactual mutations,
the properties of the mutations inferred by \texttt{sc2ts} largely follow
established results.
In Figure~\ref{fig:mutational_spectra}B we compare the mutational spectrum
in the Wide ARG to the results of \cite{Yi2021-sc},
who reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and classified the mutations occurring along the phylogeny.
We categorised all single nucleotide
mutations in the Wide ARG by type (defined by the inherited and derived states),
excluded mutations inherited by only a single sample (which are
enriched for sequencing errors and artefacts).
Similarly, we took the data for single nucleotide mutations from
\citep[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluded
mutations occurring along terminal branches, and tallied them up by type.
Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
Wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U
mutations occur more frequently than U-to-C and U-to-G, respectively. Similar
results are obtained when including the mutations inherited by only a single
sample or those occurring on terminal branches (data not shown).

@@

The Wide ARG contains 15 of these 16 recombinant sequences
(sample MILK-103C712 was removed during preprocessing; see
Section~\ref{sec:data_preprocessing}), and the results are summarised
in Tables~\ref{tab:jackson} and \ref{tab:jackson_supplement}.
Table~\ref{tab:jackson} shows the groups of sequences identified by Jackson
et al.\ as likely independent recombination events with onward transmission.
In each case we have a corresponding recombination node in the Wide ARG,
from which all the sequences in the group descend. The parent lineages
and breakpoint intervals agree closely (see
Section~\ref{sec:breakpoint_intervals} for more details on breakpoint
intervals).
For groups B, C and D,
these recombination nodes form clades consisting only of the identified
sequences.
Group A sequences were subsequently given the Pango XA designation
following onward transmission,
and there are 44 XA designated samples in the Wide ARG (including the
4 sequences analysed by Jackson et al.). The Group A recombination
node forms a monophyletic clade of these 44 samples.
Table~\ref{tab:jackson_supplement} shows the details for each of the
16 sequences individually, and showing generally a strong concordance
in mosaic structure and parent lineages
(including CAMC-CB7AB3 which is inferred to have two breakpoints under both
methods).

The Long ARG contains 5 of the sequences: two each from groups A and B
and QEUH-1067DEF. These cluster under three recombination nodes, as expected,
and have identical breakpoint intervals and parental lineage assignments
to those of Wide ARG.
The recombination nodes for Group B and QEUH-1067DEF are ancestral only
to the sequences involved.
The recombination node for group A forms a monophyletic clade of all
5 XA samples present in the Long ARG
(Figure~\ref{fig:pango-simple-origin-graph}A)

@@

When using Nextclade Pango lineage designations, the
Long ARG contains 749 samples from 33 recombinant lineages; this
reduces toe 711 and XX when singleton recombinants (i.e., recombination
nodes subtending only one sample) are removed.
In contrast, when using GISAID Pango lineage assignments (for the same
data) we have 515 samples from 38 Pango recombinant lineages,
indicating that there is substantial uncertainty in the assignment of
Pango X lineages to samples.

>>>>

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
of the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genetic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
% Early in the SARS-CoV-2 pandemic there was insufficient
% genetic variation for meaningful recombination.
Early on in the COVID-19 pandemic, the levels of genetic diversity
were too low to precipitate the emergence of distinctive recombinant strains.
By late 2020, however, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur,
and recombinant strains with onward transmission became detectable at
low frequencies~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of the SARS-CoV-2
virus.

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual process, subject
to biases [FIXME: have we a citation for how tedious this is? Also what
are the biases?]
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of characteristic mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
and as a result can only identify inter-lineage recombination events.
% specific to those motifs.
% % JK: I've left this out because I felt it broke the narrative up a bit, but
% % happy to put back in. (We're still citing the paper in a few places).
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
mutations.
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination), and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated genealogy.

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analyses, such as inference of
selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
no standard method for incorporating the effects of recombination into
phylogenetic analysis.
Ancestral Recombination Graphs (ARGs) are a means of describing such
network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
until recently lacked software support and sufficiently scalable
inference methods to be of practical use.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. The ``succinct tree sequence'' is an ARG data structure
which has led to significant computational advances across a range
of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
Wohns2022-th}, and the supporting \texttt{tskit} software library
is now widely used in population genetics applications.
The methods in \texttt{tskit} have been developed to support millions of
whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
to representing large SARS-CoV-2 genealogies,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}.
See Section~\ref{sec:args} for more details on ARGs and the succinct
tree sequence data structure.

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
Building on the open-source \texttt{tskit} library, the method infers
a joint model of mutation, genetic inheritance and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
% First, likely attachment nodes in the current ARG
% (or paths, if inferred to be a recombinant) are
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
% chosen for each of the samples in the daily batch using the
Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
in the batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of an accurate joint model of
mutation, genetic inheritance and recombination, will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts}, and to investigate the properties of the
inferred ARGs to inform subsequent development.
We work with a
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited, and
includes all 1.27 million sample genomes collected up to June 30, 2021
with a maximum delay between sampling and submission dates of 30 days
(see Section~\ref{sec:filtering_time_travellers}).
% The Long ARG is more sparsely sampled but
% includes data for an additional year (June 30, 2022).
% For the Long ARG, we use at most 1,000 randomly subsampled
% genomes per day from those meeting the maximum sampling delay criterion.
For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
day (again restricting the delay between sampling and submission to 30 days),
and include an additional year worth of samples (to June 30, 2022).

@@

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For example, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by the metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks~\citep{Kluyver2016-jupyter}
on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine in every generation,
in SARS-CoV-2 recombination is relatively rare, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes, e.g.\ restricted by time or limited to close
relatives. We expect ARGs to be particularly tree-like
early in the pandemic, when co-infection was less likely and divergence between
lineages relatively low.

@@

A classic tree-based summary of SARS-CoV-2 ancestry is provided
by the Nextstrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded Nextstrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the \texttt{sc2ts} ARGs to a backbone containing only samples present
the Nextstrain tree. This results in a small set of shared
samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
assigned to Pango recombinant lineages by Nextclade (see Section
\ref{sec:pango_x_lineages}).

@@

Figure~\ref{fig:cophylogeny} also reveals some notable differences between the
trees. Firstly, the \texttt{sc2ts} tree is generally less well resolved,
particularly in early 2020 when sampling density was much lower than
later in the pandemic.
Secondly, there are non-identical sample partitions
near the tips (e.g., in the Alpha clade).
Finally, and most noticeably, some branch lengths differ substantially.
As discussed in Section~\ref{sec:node_dating}, the dating of nodes
other than samples in \texttt{sc2ts} is currently quite crude, and
it is likely that branches not ``anchored'' by samples will be
poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
these issues may be alleviated in future versions of \texttt{sc2ts}.

@@

\subsection{Mutational spectrum}
\label{sec:mutation_spectrum}
The ARGs inferred by \texttt{sc2ts} and represented using the \texttt{tskit}
library (Section~\ref{sec:args}) are a joint estimate of the processes
of genetic inheritance, recombination and mutation. Unlike most approaches to
phylogenetic analysis, mutations are included in the \texttt{tskit}
data model alongside
the topological representation of genetic inheritance.
This has many
advantages, for example allowing us to compute statistics of the observed
sequences efficiently~\citep{Kelleher2016-wk,Ralph2020-efficiently} and
to provide high levels of data compression~\citep{Kelleher2019-ba}.
The same idea has recently been used to represent
SARS-CoV-2 data in UShER's ``mutation annotated tree''
format~\citep{Turakhia2021-ur}.

@@

Despite the presence of some artefactual mutations,
the properties of the mutations inferred by \texttt{sc2ts} largely follow
established results.
In Figure~\ref{fig:mutational_spectra}B we compare the mutational spectrum
in the Wide ARG to the results of \cite{Yi2021-sc},
who reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and classified the mutations occurring along the phylogeny.
We categorised all single nucleotide
mutations in the Wide ARG by type (defined by the inherited and derived states),
excluding mutations inherited by only a single sample (which are
more likely to be sequencing errors).
Similarly, we took the data for single nucleotide mutations from
\citet[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluding
mutations occurring along terminal branches, and tallied them up by type.
Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
Wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U
mutations occur more frequently than U-to-C and U-to-G, respectively. Similar
results are obtained when including the mutations inherited by only a single
sample or those occurring on terminal branches (data not shown).

@@

The Wide ARG contains 15 of these 16 recombinant sequences
(sample MILK-103C712 was removed during preprocessing; see
Section~\ref{sec:data_preprocessing}), and the results are summarised
in Tables~\ref{tab:jackson} and \ref{tab:jackson_supplement}.
Table~\ref{tab:jackson} shows the groups of sequences identified by Jackson
et al.\ as likely independent recombination events with onward transmission.
In each case we have a corresponding recombination node in the Wide ARG,
from which all the sequences in the group descend. The parent lineages
and breakpoint intervals agree closely (see
Section~\ref{sec:breakpoint_intervals} for more details on breakpoint
intervals).
For groups B, C and D,
these recombination nodes form clades consisting only of the identified
sequences.
Group A sequences were subsequently given the Pango XA designation
following onward transmission,
and there are 44 XA designated samples in the Wide ARG (including the
4 sequences analysed by Jackson et al.). The Group A recombination
node forms a monophyletic clade of these 44 samples.
Table~\ref{tab:jackson_supplement} shows the details for each of the
16 sequences individually, and showing generally a strong concordance
in mosaic structure and parent lineages
(including sample CAMC-CB7AB3, which is inferred to have two breakpoints under both
methods).

The Long ARG contains 5 of the sequences: two each from groups A and B
and sample QEUH-1067DEF. These cluster under three recombination nodes, as expected,
and have identical breakpoint intervals and parental lineage assignments
to those of Wide ARG.
The recombination nodes for Group B and sample QEUH-1067DEF are ancestral only
to the sequences involved.
The recombination node for group A forms a monophyletic clade of all
5 XA samples present in the Long ARG
(Figure~\ref{fig:pango-simple-origin-graph}A)

@@

When using Nextclade Pango lineage designations, the
Long ARG contains 749 samples from 33 recombinant lineages; this
reduces to 711 and FIXMEXX when singleton recombinants (i.e., recombination
nodes subtending only one sample) are removed.
In contrast, when using GISAID Pango lineage assignments (for the same
data) we have 515 samples from 38 Pango recombinant lineages,
indicating that there is substantial uncertainty in the assignment of
Pango X lineages to samples.
